---
title: "h5n1_segment_dms_summary"
output: html_document
date: '`r format(Sys.Date(), "%d %B %Y")`'
---

This file deals with mutations called against the DMS references, rather than against ancestral sequences.

# Setup
```{r setup, include=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggrepel)
library(stringr)
library(ggpubr)
library(ggpmisc)
library(lubridate)
library(ape)
library(ggtree)
library(phangorn)
library(patchwork)
library(ggnewscale)

theme_set(theme_bw())
setwd('/Users/james/Documents/convergent')
```

```{r read_mutations_data, include=F, echo=F}
mutations = read.csv('data/dms_mutations_wrt_ref_avian_ha.tsv', sep = '\t') %>%
  mutate(host = 'avian', gene = 'HA') %>%
  bind_rows(
    read.csv('data/dms_mutations_wrt_ref_bovine_ha.tsv', sep = "\t") %>%
      mutate(host = 'bovine', gene = 'HA')
  ) %>% 
  bind_rows(
    read.csv('data/dms_mutations_wrt_ref_avian_pb2.tsv', sep = "\t") %>%
      mutate(host = 'avian', gene = 'PB2')
  ) %>% 
  bind_rows(
    read.csv('data/dms_mutations_wrt_ref_bovine_pb2.tsv', sep = "\t") %>%
      mutate(host = 'bovine', gene = 'PB2')
  ) %>% 
  rename(
    dms_name = phenotype_metric_name,
    dms_value = value,
    pos_aa = position_aa
  ) %>% 
  mutate(
    date = ymd(date)
  ) %>% 
  filter(dms_name != 'mature_h5_site')
```

```{r set_up_ancestral_mutation_data, include=F, echo=F}
# read ancestral mutations
ancestor_muts_dms = read.csv('out_bovine_b3.13_ha/mutations_wrt_ref.tsv', sep = "\t") %>% 
  mutate(host='bovine', gene = 'HA') %>% 
  bind_rows(
    read.csv('out_avian_b3.13_ha/mutations_wrt_ref.tsv', sep = "\t") %>% 
    mutate(host='avian', gene = 'HA')
  ) %>% 
  bind_rows(
    read.csv('out_bovine_b3.13_pb2/mutations_wrt_ref.tsv', sep = "\t") %>% 
    mutate(host='bovine', gene = 'PB2')
  ) %>% 
  bind_rows(
    read.csv('out_avian_b3.13_pb2/mutations_wrt_ref.tsv', sep = "\t") %>% 
    mutate(host='avian', gene = 'PB2')
  ) %>% 
  rename(accession = node) %>% 
  filter(str_starts(accession, 'NODE_'))

# read dates for ancestors
ancestor_dates = read.csv('out_avian_b3.13_ha/treetime/dates.tsv', sep='\t') %>% 
  mutate(host='avian', gene='HA') %>% 
  select(! c(numeric.date)) %>% 
  bind_rows(
    read.csv('out_bovine_b3.13_ha/treetime/dates.tsv', sep='\t') %>% 
      mutate(host='bovine', gene='HA') %>% 
      select(! c(numeric.date))
  ) %>% 
   bind_rows(
    read.csv('out_avian_b3.13_pb2/treetime/dates.tsv', sep='\t') %>% 
      mutate(host='avian', gene='PB2') %>% 
      select(! c(numeric.date))
  ) %>%
   bind_rows(
    read.csv('out_bovine_b3.13_pb2/treetime/dates.tsv', sep='\t') %>% 
      mutate(host='bovine', gene='PB2') %>% 
      select(! c(numeric.date))
  ) %>% 
  rename(accession = X.node) %>% 
  mutate(date = ymd(date)) %>% 
  filter(str_starts(accession, 'NODE_'))
  

# join in dates
ancestor_muts_dms = left_join(
  ancestor_muts_dms,
  ancestor_dates,
  by = join_by(accession, host, gene)
)

# read dms data
dms = read.csv('data/dms_data_ha.tsv', sep = '\t') %>%
  rename(pos_aa = position_aa,
         dms_name = name,
         dms_value = value,) %>%
  mutate(gene = 'HA') %>%
  select(!c('ref_codon', 'alt_codon', 'gff_feature')) %>%
  filter(!dms_name %in% c('mature_h5_site')) %>%
  distinct() %>%
  bind_rows(
    read.csv('data/summary_prefs_effects_diffsel.csv') %>%
      select(c('site', 'wildtype', 'mutation', 'mutdiffsel')) %>%
      rename(
        pos_aa = site,
        ref_aa = wildtype,
        alt_aa = mutation
      ) %>%
      mutate(gene = 'PB2') %>%
      pivot_longer(
        cols = mutdiffsel,
        names_to = 'dms_name',
        values_to = 'dms_value'
      ) %>%
      filter(!is.na(dms_value))
  )

# join in dms data
ancestor_muts_dms = ancestor_muts_dms %>% 
  left_join(
    dms,
    by = join_by(gene, pos_aa, ref_aa, alt_aa),
    relationship = "many-to-many"
  )

```


```{r sum_dms_values, include=F, echo=F}
mutations_summed_dms =
  mutations %>%
  summarize(
    dms_value_sum = sum(dms_value),
    n_dms_values = n(),
    .by = c(accession, host, date, dms_name, gene)
  ) %>%
  mutate(seq_type = 'sample') %>%
  bind_rows(
    ancestor_muts_dms %>%
      filter(!is.na(dms_name)) %>% 
      summarize(
        dms_value_sum = sum(dms_value),
        n_dms_values = n(),
        .by = c(accession, date, host, dms_name, gene)
      ) %>% 
      mutate(seq_type = 'ancestral')
  )

mutations_summed_dms_cows_vs_all = read.csv('data/mutations_dms_non_bovine_non_b3_13.tsv', sep = '\t') %>% 
  rename(
    date = mid_collection_date,
    host_raw = host,
    dms_name = phenotype_metric_name,
    gene = gff_feature
  ) %>% 
  filter(dms_name != 'mature_h5_site') %>% 
  mutate(
    gene = case_when(
      gene == 'XAJ25426.1' ~ 'PB2', 
      gene == 'XAJ25415.1' ~ "HA",
      .default = gene
    ),
    host = 'non-bovine',
    date = ymd(date)
  ) %>% 
  bind_rows(
    mutations_summed_dms %>% 
      filter(host == 'bovine') %>% 
      mutate(
        lineage_name = 'B3.13'
      ) %>% 
      select(!seq_type)
  )
```

## QC Plots

```{r count_dms_values_summed, include=T, echo=F, fig.width=14, fig.height=9}
# How many values represent each thing
mutations_summed_dms %>% filter(seq_type == 'sample') %>% 
  ggplot(aes(x=n_dms_values)) +
  geom_bar() +
  facet_wrap(vars(dms_name, host, gene), labeller = label_wrap_gen(multi_line=FALSE), scales = 'free_y') +
  ggtitle("count dms values summed: samples")

mutations_summed_dms %>% filter(seq_type == 'ancestral') %>% 
  ggplot(aes(x=n_dms_values)) +
  geom_bar() +
  facet_wrap(vars(dms_name, host, gene), labeller = label_wrap_gen(multi_line=FALSE), scales = 'free_y') +
  ggtitle("count dms values summed: ancestral nodes")

mutations_summed_dms_cows_vs_all %>% filter(host == 'non-bovine') %>% 
  ggplot(aes(x=n_dms_values)) +
  geom_bar() +
  facet_wrap(vars(dms_name, host, gene), labeller = label_wrap_gen(multi_line=FALSE), scales = 'free_y') +
  ggtitle("count dms values summed: non-bovine, non-B3.13 samples")

mutations_summed_dms_cows_vs_all %>% 
  select(accession, date) %>% 
  distinct() %>% 
  ggplot() +
  geom_histogram(aes(x=date, y=after_stat(density)), bins=60) +
  ggtitle('Distribution of sample dates')
```

### Missing DMS Values: ancestral seqs

How many of our mutations do we have DMS data for?
This is counting by position, ref, and alt. 
This is only for ancestral sequence mutations (wrt ref).

```{r missing_dms_values, include=T, echo=F}
# count how many positions are missing values
wide_dms_by_mutation = ancestor_muts_dms %>% 
  filter(alt_aa != ref_aa) %>% 
  pivot_wider(
    names_from=c(dms_name),
    values_from=dms_value,
    names_prefix='dms_'
  ) %>% 
  # shim is na for all, which lets us get the totals easily
  mutate(dms_shim_const_NA=NA)

calc_fracs_present = function(df){
  return(
    df %>%  
    mutate(across(starts_with('dms_'), ~ 1 - (.x/dms_shim_const_NA), .names='frac_present_{.col}')) %>%
    select(!(starts_with('dms_') | contains('NA', ignore.case=F))) %>% 
    pivot_longer(names_to='dms_name', values_to='frac_present', cols=starts_with('frac_')) %>% 
    mutate(dms_name = str_remove(dms_name, 'frac_present_dms_')) 
    # pivot_wider(names_from = c(host), values_from=frac_present, names_prefix = 'frac_present_')
  )
}

dms_values_present_by_mutation = wide_dms_by_mutation %>% 
  summarize(across(starts_with('dms_'), ~ sum(is.na(.x))), .by=c(host, gene)) %>% 
  calc_fracs_present()

ggplot(dms_values_present_by_mutation, aes(x=dms_name, y=frac_present)) +
  geom_col(aes(fill=interaction(host, gene)), position=position_dodge()) +
  theme(axis.text.x = element_text(angle = -30, vjust = 0.5, hjust=0))
```

# Overall distributions
## Samples

```{r dist_of_summed_values, include=T, echo=F, fig.width=13, fig.height=10}
mutations_summed_dms %>% filter(seq_type == 'sample') %>% 
ggplot(aes(x = dms_value_sum)) +
  geom_histogram() +
  facet_wrap(vars(dms_name, host, gene), labeller = label_wrap_gen(multi_line=FALSE), scales='free', ncol=4) +
  ggtitle('distribution of dms sums within sample seqs')

mutations_summed_dms %>% filter(seq_type == 'sample') %>% 
ggplot(aes(x = date, y=dms_value_sum)) +
  geom_point(aes(color = n_dms_values)) +
  scale_color_viridis_c() +
  facet_wrap(vars(dms_name, host, gene), labeller = label_wrap_gen(multi_line=FALSE), scales='free_y', ncol=4) +
  ggtitle('dms sums by date within sample seqs')

```

## Ancestors
```{r dist_of_summed_values_ancestors, include=T, echo=F, fig.width=13, fig.height=10}
mutations_summed_dms %>% filter(seq_type == 'ancestral') %>% 
ggplot(aes(x = dms_value_sum)) +
  geom_histogram() +
  facet_wrap(vars(dms_name, host, gene), labeller = label_wrap_gen(multi_line=FALSE), scales='free', ncol=4) +
  ggtitle('distribution of dms sums within ancestral seqs')

mutations_summed_dms %>% filter(seq_type == 'ancestral') %>% 
ggplot(aes(x = date, y=dms_value_sum)) +
  geom_point(aes(color = n_dms_values)) +
  scale_color_viridis_c() +
  facet_wrap(vars(dms_name, host, gene), labeller = label_wrap_gen(multi_line=FALSE), scales='free_y', ncol=4) +
    ggtitle('dms sums by date within ancestral seqs')
```

## Cows vs the world

The "Cows" here are the bovine B3.13 samples used in the rest of my analysis.
The "world" is a set of samples from non-bovine hosts, with lineages other than
B3.13 listed (so any sample with no lineage is excluded) and with a "collection span" <= 10 days.
See the SQL queries for details.

```{r dist_of_summed_values_cows_vs_world, include=T, echo=F, fig.width=13, fig.height=11}
pl = plot_list()
for (name in (mutations_summed_dms_cows_vs_all %>% select(dms_name) %>% unique())$dms_name) {
 pl = pl + mutations_summed_dms_cows_vs_all %>% filter(dms_name == name) %>% 
  ggplot(aes(x = dms_value_sum)) +
    geom_histogram(aes(y=after_stat(density)), bins=50) +
    facet_wrap(vars(host, gene), labeller = label_wrap_gen(multi_line=FALSE), scales='fixed', ncol=1) +
    ggtitle(name) +
    theme(plot.title = element_text(size = 10, vjust = 0))
}
pl = pl + plot_layout(ncol=3) +
  plot_annotation(
    title = 'DMS value sum distributions, bovine B3.13 vs non-bovine non-B3.13' 
  )
print(pl)  
  
mutations_summed_dms_cows_vs_all %>% 
ggplot(aes(x = date, y=dms_value_sum)) +
  geom_point(aes(color = n_dms_values)) +
  scale_color_viridis_c() +
  facet_wrap(vars(dms_name, host, gene), labeller = label_wrap_gen(multi_line=FALSE), scales='free_y', ncol=4) +
  ggtitle('dms sums by date, bovine B3.13 vs non-bovine non-B3.13')
```

```{r non-bovine_only, include=T, echo=F, fig.height=15, fig.width=15}
mutations_summed_dms_cows_vs_all %>% filter(host == 'non-bovine') %>%
  inner_join(
    mutations_summed_dms_cows_vs_all %>%
      filter(host == 'non-bovine') %>%
      summarize(count_samples = n(), .by = host_raw),
    by = join_by(host_raw)
  ) %>%
  filter(count_samples >= 100) %>%
  ggplot(aes(x=host_raw, y=dms_value_sum)) +
  geom_boxplot() +
  facet_wrap(vars(dms_name, gene), labeller = label_wrap_gen(multi_line=FALSE), scales='free_y', ncol=2) +
  theme(axis.text.x = element_text(angle = -90, vjust = 0.5, hjust=0, size = 8)) +
  ggtitle('dms value sums for non-B3.13, non-bovine hosts with >= 100 samples')
```


<!-- # Go to Ombrosa, take to the trees -->
# Trees

```{r set_up_trees, include=F, echo=F}

tree_bovine_ha = read.nexus('out_bovine_b3.13_ha/treetime/timetree.nexus')
tree_avian_ha = read.nexus('out_avian_b3.13_ha/treetime/timetree.nexus')
tree_bovine_pb2 = read.nexus('out_bovine_b3.13_pb2/treetime/timetree.nexus')
tree_avian_pb2 = read.nexus('out_avian_b3.13_pb2/treetime/timetree.nexus')

get_dms_values = function(nodes, dms_name) {
  joined = data.frame(accession=nodes, dms_name=dms_name) %>% 
    left_join(mutations_summed_dms, by=join_by(accession, dms_name)) %>% 
    select(accession, dms_value_sum)
  return(joined)
}

```

```{r calc_deltas_parent_to_child, include=F, echo=F}
get_parent_child_deltas = function(tree, host_, gene_) {
  edge_dms = data.frame()
  edges = tree$edge %>% 
      as.data.frame() %>% 
      mutate(index = row_number())
  
  all_labels = c(tree$tip.label, tree$node.label)
  for(child in all_labels) {
    parent = all_labels[Ancestors(tree, child, type='parent')]
    if(length(parent) == 0){
      next
    }
    child_idx = which(all_labels == child)
    parent_idx = which(all_labels == parent)
    edge_idx = edges %>% filter(V1 == parent_idx & V2 == child_idx) %>% pull(index)
    
    edge_dms = edge_dms %>% bind_rows(
      mutations_summed_dms %>% 
        filter(gene == gene_ & host == host_ & accession == child) %>%
        inner_join(
          mutations_summed_dms %>% 
              filter(gene == gene_ & host == host_ & accession == parent),
          by = join_by(gene, host, dms_name),
          suffix = c("_child", "_parent")
        ) %>% 
        mutate(edge_index = edge_idx)
    )
  }
  edge_dms = edge_dms %>% 
    mutate(
      dms_value_sum_delta = dms_value_sum_child - dms_value_sum_parent
    )
  return(edge_dms)
}

get_edge_dms_deltas_vec = function(tree, edge_dms, dms_name) {
  n_edges = nrow(tree$edge)
  scaffold = data.frame(edge_index = seq(n_edges), dms_name = dms_name)
  edge_deltas = (
    scaffold %>% 
      left_join(
        edge_dms,
        by = join_by(edge_index, dms_name)
      )
  )$dms_value_sum_delta
  edge_deltas = c(NA, edge_deltas)
  return(edge_deltas)
}

```

## Bovine HA

```{r plot_dms_colored_trees_bovine_ha, include=T, echo=F, fig.height=20, fig.width=10}

plot_dms_tree = function(dms_name, tree, host, gene){
  dms_values = get_dms_values(tree$tip.label, dms_name)
  p = ggtree(tree)
  p = p %<+% dms_values +
      geom_tippoint(aes(color=dms_value_sum), size=2) +
      scale_color_viridis_c() +
      ggtitle(paste(paste(host, gene, "B3.13", sep = ', '), dms_name, sep = " | ")) +
      theme(plot.title = element_text(size = 16, face = "bold", vjust = 0)) +
      theme(panel.border = element_rect()) +
      expand_limits(y = c(-20, length(tree$tip.label) + 20))
  print(p)
}

for (name in (mutations_summed_dms %>% filter(gene == 'HA') %>% select(dms_name) %>% unique())$dms_name) {
  plot_dms_tree(name, tree_bovine_ha, "Bovine", "HA")
}
```

## Avian HA

```{r plot_dms_colored_trees_avian_ha, include=T, echo=F, fig.height=20, fig.width=10}
for (name in (mutations_summed_dms %>% filter(gene == 'HA') %>% select(dms_name) %>% unique())$dms_name){
  plot_dms_tree(name, tree_avian_ha, 'Avian', "HA")
}
```

## PB2

```{r plot_dms_colored_trees_pb2, include=T, echo=F, fig.height=20, fig.width=10}
plot_dms_tree("mutdiffsel", tree_bovine_pb2, "Bovine", "PB2")
plot_dms_tree("mutdiffsel", tree_avian_pb2, "Avian", "PB2")
```

## Ancestors

```{r plot_dms_colored_trees_ancestors, include=T, echo=F, fig.height=20, fig.width=10}
plot_dms_tree_ancestors = function(dms_name, tree, host, gene){
  dms_values = get_dms_values(tree$node.label, dms_name)
  p = ggtree(tree)
  p = p %<+% dms_values +
      geom_nodepoint(aes(color=dms_value_sum), size=2) +
      scale_color_viridis_c() +
      ggtitle(paste(paste(host, gene, "B3.13", sep = ', '), dms_name, sep = " | ")) +
      theme(plot.title = element_text(size = 16, face = "bold", vjust = 0)) + 
      theme(panel.border = element_rect()) +
      expand_limits(y = c(-20, length(tree$tip.label) + 20))
  print(p)
}

for (name in (mutations_summed_dms %>% filter(gene == 'HA') %>% select(dms_name) %>% unique())$dms_name) {
  plot_dms_tree_ancestors(name, tree_bovine_ha, "Bovine", "HA")
}

for (name in (mutations_summed_dms %>% filter(gene == 'HA') %>% select(dms_name) %>% unique())$dms_name) {
  plot_dms_tree_ancestors(name, tree_avian_ha, "Avian", "HA")
}

plot_dms_tree_ancestors("mutdiffsel", tree_bovine_pb2, "Bovine", "PB2")
plot_dms_tree_ancestors("mutdiffsel", tree_avian_pb2, "Avian", "PB2")
```

### Delta parent-to-child

Here, edges are colored by the change in summed dms value between parent and child.
That is: `child_value - parent_value`.
These have not been very useful, so disabled for now. 

```{r plot_parent_child_deltas, include=T, echo=F, fig.height=20, fig.width=10, eval=F}
plot_dms_tree_edge_deltas = function(tree, edge_dms, host, gene, dms_name){
  edge_deltas_vec = get_edge_dms_deltas_vec(tree, edge_dms, dms_name)

    p = ggtree(tree, aes(color=edge_deltas_vec)) +
      scale_color_viridis_c() +
      ggtitle(paste(paste(host, gene, "B3.13", sep = ', '), dms_name, sep = " | ")) +
      labs(color='child - parent') +
      theme(plot.title = element_text(size = 16, face = "bold", vjust = 0)) + 
      theme(panel.border = element_rect()) +
      expand_limits(y = c(-20, length(tree$tip.label) + 20))
  print(p)
}

edge_dms_bovine_ha = get_parent_child_deltas(tree_bovine_ha, 'bovine', 'HA')
for (name in (mutations_summed_dms %>% filter(gene == 'HA') %>% select(dms_name) %>% unique())$dms_name){
  plot_dms_tree_edge_deltas(tree_bovine_ha, edge_dms_bovine_ha, 'bovine', 'HA', name)
}



edge_dms_avian_ha = get_parent_child_deltas(tree_avian_ha, 'avian', 'HA')
for (name in (mutations_summed_dms %>% filter(gene == 'HA') %>% select(dms_name) %>% unique())$dms_name){
  plot_dms_tree_edge_deltas(tree_avian_ha, edge_dms_avian_ha, 'avian', 'HA', name)
}

edge_dms_bovine_pb2 = get_parent_child_deltas(tree_bovine_pb2, 'bovine', 'PB2')
plot_dms_tree_edge_deltas(tree_bovine_pb2, edge_dms_bovine_pb2, 'bovine', 'PB2', 'mutdiffsel')

edge_dms_avian_pb2 = get_parent_child_deltas(tree_avian_pb2, 'avian', 'PB2')
plot_dms_tree_edge_deltas(tree_avian_pb2, edge_dms_avian_pb2, 'avian', 'PB2', 'mutdiffsel')

```

# DMS time series

```{r calc_branch_dms_time_series, include=F, echo=F}
get_branch_dms_ts = function(tree, host_, gene_) {
  
  branch_dms = data.frame()
  
  # tree has two vectors of labels, tip.label and node.label
  # Ancestors() treats them as numbered 1, ..., N, N+1, N+M
  # where N is the number of samples, and M is the number of internal nodes.
  all_labels = c(tree$tip.label, tree$node.label)
  for(sample in tree$tip.label) {
    ancestor_idxs = Ancestors(tree, sample)
    ancestors = all_labels[ancestor_idxs]
    
    branch_dms = branch_dms %>% bind_rows(
      data.frame(
        accession = c(sample, ancestors),
        branch_leaf = sample,
        ancestor_number = seq(0, length(ancestors)),
        ancestor_count = length(ancestors),
        gene = gene_,
        host = host_
      )
    )
  }
  
  branch_dms = left_join(
    branch_dms,
    mutations_summed_dms,
    by = join_by(accession, gene, host),
    relationship = 'many-to-many'
  ) %>% 
    filter(!is.na(dms_name)) %>%
    group_by(branch_leaf, dms_name) %>%
    mutate(
      delta_dms_sum = dms_value_sum - lag(dms_value_sum, order_by = -ancestor_number)
    )
        
  return(branch_dms)
}

branches_dms_bovine_ha = get_branch_dms_ts(tree_bovine_ha, 'bovine', 'HA')
branches_dms_bovine_pb2 = get_branch_dms_ts(tree_bovine_pb2, 'bovine', 'PB2')
```


```{r plot_branch_dms_ts, include=T, echo=F, fig.height=7, fig.width=10}
# this is selecting the clade that shows higher sa26 usage increase.
# leaf_sample = tree_bovine_ha$tip.label[
#     Descendants(tree_bovine_ha, 'NODE_0001662', type='tips') %>% 
#       unlist()
#   ]
# print(length(leaf_sample))

# tried using as the time axis:
# frac_along_branch = (ancestor_count - ancestor_number) / ancestor_count
# didn't really help... (31 mar 2026)
# maybe something like this would work better if it were based on the dates instead?

# branches_dms_bovine_ha %>% 
#   filter(branch_leaf %in% leaf_sample) %>% 
#   filter(dms_name == 'sa26_usage_increase') %>% 
#   ggplot(aes(x=date, y=dms_value_sum)) +
#   geom_line(aes(group = branch_leaf, color=branch_leaf), alpha=0.4) +
#   scale_color_discrete(guide='none') +
#   facet_wrap(vars(dms_name), labeller = label_wrap_gen(multi_line=FALSE), scales='free_y', ncol=2) +
#   ggtitle('DMS values along branch | Bovine HA')
# 
# branches_dms_bovine_ha %>% 
#   filter(branch_leaf %in% leaf_sample) %>% 
#   filter(dms_name == 'sa26_usage_increase') %>%
# ggplot(aes(x=date, y=delta_dms_sum)) +
#   geom_line(aes(group = branch_leaf, color=branch_leaf, alpha=0.4), alpha=1, na.rm=T) +
#   scale_color_discrete(guide='none') +
#   facet_wrap(vars(dms_name), labeller = label_wrap_gen(multi_line=FALSE), scales='free_y', ncol=2) +
#   ggtitle('Delta in DMS value at each node in branch | Bovine HA')

# branches_dms_bovine_ha %>% 
#   filter(branch_leaf %in% leaf_sample) %>% 
#   filter(dms_name == 'sa26_usage_increase') %>% 
# ggplot(aes(x=date, y=cum_delta_dms_sum)) +
#   geom_line(aes(group = branch_leaf, color=branch_leaf), alpha=0.4) +
#   scale_color_discrete(guide='none') +
#   facet_wrap(vars(dms_name), labeller = label_wrap_gen(multi_line=FALSE), scales='free_y', ncol=2) +
#   ggtitle('Cumulative delta in DMS value at each node in branch | Bovine HA')

leaf_sample = c('SRR33943324', 'SRR32633084', 'SRR29694479', 'SRR30811300', 'SRR31820877', 'SRR31820835', 'SRR34270150', 'SRR30090365', 'SRR31820843', 'SRR32868633', 'SRR29543338', 'SRR32125668', 'SRR33124614', 'SRR31960125', 'SRR31820958', 'SRR31820946', 'SRR32125732', 'SRR32254306', 'SRR29740713', 'SRR32006798')

branches_dms_bovine_ha %>% 
  filter(branch_leaf %in% leaf_sample) %>% 
  ggplot(aes(x=date, y=dms_value_sum)) +
  geom_line(aes(group = branch_leaf, color=branch_leaf), alpha=0.4) +
  # scale_color_discrete(guide='none') +
  facet_wrap(vars(dms_name), labeller = label_wrap_gen(multi_line=FALSE), scales='free_y', ncol=2) +
  ggtitle('DMS values along branch | Bovine HA')

branches_dms_bovine_ha %>% 
  filter(branch_leaf %in% leaf_sample) %>% 
ggplot(aes(x=date, y=delta_dms_sum)) +
  geom_line(aes(group = branch_leaf, color=branch_leaf, alpha=0.4), alpha=1, na.rm=T) +
  # scale_color_discrete(guide='none') +
  facet_wrap(vars(dms_name), labeller = label_wrap_gen(multi_line=FALSE), scales='free_y', ncol=2) +
  ggtitle('Delta in DMS value at each node in branch | Bovine HA')

```

```{r dms_timeseries_regression_bovine_ha, echo=F, include=T, fig.width=12, fig.height=10}
branches_dms_bovine_ha %>% 
  ggplot(aes(x=date, y=dms_value_sum)) +
  geom_point(alpha=0.2) +
  # geom_line(aes(group = branch_leaf), alpha=0.1) +
  facet_wrap(vars(dms_name), labeller = label_wrap_gen(multi_line=FALSE), scales='free_y', ncol=3) +
  ggtitle('DMS values along branch | Bovine HA') +
  stat_poly_line() +
  stat_poly_eq(color='blue')

# also tried 'NODE_0000049', 'NODE_0000419', they were largely similar
clade_root = 'NODE_0002945' 
leaf_sample = tree_bovine_ha$tip.label[
    Descendants(tree_bovine_ha, clade_root, type='tips') %>%
      unlist()
  ]

branches_dms_bovine_ha %>%
  filter(branch_leaf %in% leaf_sample) %>%
  ggplot(aes(x=date, y=dms_value_sum)) +
  geom_point(alpha=0.2) +
  geom_line(aes(group = branch_leaf), alpha=0.1) +
  facet_wrap(vars(dms_name), labeller = label_wrap_gen(multi_line=FALSE), scales='free_y', ncol=3) +
  stat_poly_line() +
  stat_poly_eq(color='blue') +
    ggtitle(
    paste(
      'DMS values along branch', 
      paste(length(leaf_sample), 'descendants of', clade_root),  
      'Bovine HA', sep=' | '
      )
    )

```

```{r dms_timeseries_regression_bovine_pb2, echo=F, include=T, fig.width=12, fig.height=10}
branches_dms_bovine_pb2 %>% 
  ggplot(aes(x=date, y=dms_value_sum)) +
  geom_point(alpha=0.2) +
  geom_line(aes(group = branch_leaf), alpha=0.1) +
  facet_wrap(vars(dms_name), labeller = label_wrap_gen(multi_line=FALSE), scales='free_y', ncol=3) +
  ggtitle('DMS values along branch | Bovine PB2') +
  stat_poly_line() +
  stat_poly_eq(color='blue')
```

# Local branching index

LBI is calculated using the treetime divergence tree, but the trees shown here are the time trees.
They have identical topology and node names, but the divergence tree avoids mixing time into the distances between nodes.

Here are the equations, updated to reflect my understanding based on the python implementation at https://github.com/rneher/FitnessInference/blob/master/prediction_src/fitness_inference.py.

The up-message from a node $i$ to its parent is:
$$
\mathrm{up}_i = \tau \left(1 - \mathrm{exp}\left(\frac{-(b_i + b_\varepsilon)}{ts \cdot \tau}\right)\right) 
+ \mathrm{exp}\left(\frac{-(b_i + b_\varepsilon)}{ts \cdot \tau}\right) \sum_j \mathrm{up}_j
$$
Where $b_i$ is the distance from its parent to node $i$, and nodes $j$ are children of $i$.

The down-message from $i$ to its child $j$ is given by:
$$
\mathrm{dn}_{ij} = \tau \left(1 - \mathrm{exp}\left(\frac{-(b_j + b_\varepsilon)}{ts \cdot \tau}\right)\right) 
+ \mathrm{exp}\left(\frac{-(b_j + b_\varepsilon)}{ts \cdot \tau}\right) \left( \mathrm{dn}_{\cdot i} + \sum_{k \neq j} \mathrm{up}_{k} \right)
$$
Where $b_j$ is distance from $i$ to $j$, $\mathrm{dn}_{\cdot i}$ is the down-message passed to $i$ from its parent, and nodes $k$ are children of $i$.

$b_\varepsilon$ is a small shim value used to prevent having branch lengths of 0.
The paper's implementation uses both 1e-5 and 1e-7, and I used the latter in my implementation.
I'm not entirely sure what they're trying to prevent here, I don't think that allowing the branch length to be 0 would radically change the shape of the output.

$ts$ is a time scale value, calculated based on a simple pairwise distance metric on the aligned sequences.
$ts = \frac{\text{mean pairwise distance}}{2}$.

$\tau$ is a scalar used to control how large a neighborhood around each node is considered in calculating its score. 
The paper says that, optimally, $\tau = 0.0625$, but in the implementation, twice that value is used in the actual calculations.
As I understand it, this is because the value of $0.0625$ is meant to be with respect to pairwise genetic distance, but the branch lengths in the tree are instead in units of pair coalescent time, which is assumed to be related to twice the genetic distance.
In my implementation I've copied this and used a value of $\tau = 2 \cdot 0.625$.

Once all the up and down messages are in place, the final value for each node $i$ is given by:
$$
\lambda_i(\tau) = \mathrm{dn}_{\cdot i} + \sum_j \mathrm{up}_j
$$

```{r set_up_lbi, echo=F, include=F}
local_branching_index = read.csv('out_bovine_b3.13_ha/local_branching_index_shim.tsv', sep='\t') %>% 
  mutate(gene='HA', host='bovine') %>% 
  bind_rows(
    local_branching_index = read.csv('out_bovine_b3.13_pb2/local_branching_index_shim.tsv', sep='\t') %>% 
    mutate(gene='PB2', host='bovine')
  ) %>% 
  bind_rows(
    local_branching_index = read.csv('out_avian_b3.13_ha/local_branching_index_shim.tsv', sep='\t') %>% 
    mutate(gene='HA', host='avian')
  ) %>% 
  bind_rows(
    local_branching_index = read.csv('out_avian_b3.13_pb2/local_branching_index_shim.tsv', sep='\t') %>% 
  mutate(gene='PB2', host='avian')
  ) %>% 
  rename(accession = node)

mutations_summed_dms = mutations_summed_dms %>% 
  select(!matches('^lbi$')) %>% 
  left_join(
      local_branching_index,
      by = join_by(accession, host, gene),
      relationship = 'many-to-one'
    )
```

```{r plot_lbi_trees, echo=F, include=T, fig.height=20, fig.width=10}
plot_lbi_tree = function(tree, host_, gene_){
  ylim_expansion = round(0.01 * length(tree$tip.label))
  ylim = c(-ylim_expansion, length(tree$tip.label) + ylim_expansion)
  
  lbi_values = local_branching_index %>% 
    filter(gene == gene_ & host == host_) %>% 
    select(accession, lbi)
  p = ggtree(tree, color='grey')
  p = p %<+% lbi_values +
      geom_tippoint(aes(color=lbi), size=2) +
      geom_nodepoint(aes(color=lbi), size=2) +
      scale_color_viridis_c() +
      ggtitle(paste(paste(host_, gene_, "B3.13", sep = ', '), 'local branching index', sep = " | ")) +
      theme(plot.title = element_text(size = 16, face = "bold", vjust = 0)) +
      theme(panel.border = element_rect()) +
      expand_limits(y = ylim)
  print(p)
}

plot_lbi_tree(tree_bovine_ha, 'bovine', 'HA')
plot_lbi_tree(tree_avian_ha, 'avian', 'HA')
plot_lbi_tree(tree_bovine_pb2, 'bovine', 'PB2')
plot_lbi_tree(tree_avian_pb2, 'avian', 'PB2')
```

```{r lbi_vs_dms, echo=F, include=T, fig.height=10, fig.width=10}

mutations_summed_dms %>% 
  ggplot(aes(x=dms_value_sum, y=lbi)) +
  geom_point(alpha=0.2)  +
  facet_wrap(vars(host, gene, dms_name), labeller = label_wrap_gen(multi_line=FALSE), scales='free_x', ncol=4)

```

## SC2 comparison
```{r lbi_sc2_comparison, include=T, echo=F, fig.width=15, fig.height=50}
lbi_sc2 = read.csv('out_sc2_nextstrain/local_branching_index_shim.tsv', sep='\t') %>% 
  rename(accession = node)

tree_sc2 = read.nexus('out_sc2_nextstrain/nextstrain_ncov_open_global_6m_tree.nexus')

lbi_sc2 = lbi_sc2 %>% 
  left_join(
  read.csv('out_sc2_nextstrain/metadata.tsv', sep='\t') %>% 
    select(strain, Nextstrain_clade, pango_lineage) %>% 
    rename(accession = strain),
  by=join_by(accession)
)

lbi_sc2 = lbi_sc2 %>% 
  mutate(pango_short = str_extract(pango_lineage, '[A-z]+(\\.\\d+){0,1}')) 

rare_pango_shorts = lbi_sc2 %>% 
  summarize(n = n(), .by = pango_short) %>% 
  filter(n < 10) %>% 
  select(pango_short) %>% 
  unlist()
  
lbi_sc2 = lbi_sc2 %>% mutate(
    pango_short = case_when(
      pango_short %in% rare_pango_shorts ~ 'other (n < 10)',
      .default = pango_short
    )
  )

plot_lbi_tree_sc2 = function(tree, lbi_values){
  ylim_expansion = round(0.01 * length(tree$tip.label))
  ylim = c(-ylim_expansion, length(tree$tip.label) + ylim_expansion)
  
      
  p = ggtree(tree, color='grey')
  p = p %<+% lbi_values +
      geom_tippoint(aes(color=pango_short), size=1.3, shape = 15 ) +
      scale_color_discrete() +
      guides(color=guide_legend(ncol = 1)) +
      ggnewscale::new_scale_color() +
      geom_nodepoint(aes(color=lbi), size=1.3) +
      scale_color_viridis_c() +
      ggtitle("SC2 LBI") +
      theme(plot.title = element_text(size = 16, face = "bold", vjust = 0)) +
      theme(panel.border = element_rect()) +
      expand_limits(y = ylim) 

  print(p)
}
# plot_lbi_tree_sc2(tree_sc2, lbi_sc2)

```

```{r lineage_props_by_node, echo=F, include=T, fig.width=60, fig.height=80}
filter_to_descendant_tips = function(df, node) {
  tip_idxs = unlist(Descendants(tree_sc2, node, type='tips'))
  tip_labels = tree_sc2$tip.label[tip_idxs]
  return(
    df %>% filter(accession %in% tip_labels)
  )
}

count_tips_pango = function() {
  pango_counts = data.frame()
  for (node in tree_sc2$node.label) {
   
    tips = filter_to_descendant_tips(lbi_sc2, node) %>% 
      summarize(n_pango = n(), .by = pango_lineage)
    
    n_desc_tips = sum(tips$n_pango)
    
      
      tips = tips %>% 
        mutate(
          n_desc_tips = n_desc_tips,
          ancestor = node
        )
    pango_counts = pango_counts %>% bind_rows(tips)
  }
  pango_counts = pango_counts %>% 
    left_join(
      lbi_sc2 %>% 
        select(accession, lbi), 
      by = join_by(ancestor == accession)
    ) %>% 
    rename(ancestor_lbi = lbi)
  return(pango_counts)
}

count_tips_pango_short = function() {
  pango_counts = data.frame()
  for (node in tree_sc2$node.label) {
   
    tips = filter_to_descendant_tips(lbi_sc2, node) %>% 
      summarize(n_pango_short = n(), .by = pango_short)
    
    n_desc_tips = sum(tips$n_pango_short)
    
      
      tips = tips %>% 
        mutate(
          n_desc_tips = n_desc_tips,
          ancestor = node
        )
    pango_counts = pango_counts %>% bind_rows(tips)
  }
  pango_counts = pango_counts %>% 
    left_join(
      lbi_sc2 %>% 
        select(accession, lbi), 
      by = join_by(ancestor == accession)
    ) %>% 
    rename(ancestor_lbi = lbi)
  return(pango_counts)
}

pango_counts = count_tips_pango()

pango_counts %>%
  filter(n_desc_tips >= 1000) %>% 
  ggplot(aes(x = ancestor, y=n_pango/n_desc_tips)) +
  geom_col(aes(fill=pango_lineage, group=ancestor), show.legend = F) +
  guides()

pango_counts %>%
  filter(n_desc_tips >= 1000) %>% 
  ggplot(aes(x=ancestor_lbi, y=n_pango/n_desc_tips)) +
  geom_point(aes(color = pango_lineage), show.legend = T, alpha=0.2)

median_lbi = pango_counts %>% 
  select(ancestor, ancestor_lbi) %>% 
  distinct() %>% 
  summarize(mlbi = median(ancestor_lbi)) %>% 
  pull

pango_counts %>% 
  mutate(pango_frac = n_pango/n_desc_tips) %>% 
  pivot_wider(names_from = pango_lineage, values_from = pango_frac, names_prefix = 'frac_', values_fill = 0) %>%
  pivot_longer(cols = starts_with('frac_'), names_to = 'pango_lineage', names_prefix = 'frac_', values_to = 'pango_frac') %>%
  arrange(ancestor_lbi) %>% 
  ggplot(aes(y=pango_lineage, x=ancestor)) +
  geom_tile(aes(fill=pango_frac)) +
  scale_fill_continuous(type='viridis') +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size=16)) +
  theme(axis.text.y = element_text(size=10)) +
  scale_y_discrete(sec.axis = dup_axis()) +
  scale_x_discrete(
    labels = function(s){
      lbis = pango_counts %>% 
        arrange(ancestor_lbi) %>% 
        filter(ancestor %in% s) %>% 
        distinct(ancestor, ancestor_lbi) %>% 
        select(ancestor_lbi) %>% 
        pull %>% 
        round(3)
      for (i in seq(length(lbis))){
        if (i %% 20 != 0) {
          lbis[i] = ''
        }
      }
      return(lbis)
    }
  )

```

```{r, include=T, echo=F, fig.width=60, fig.height=40}
pango_short_counts = count_tips_pango_short()

node_sample = pango_short_counts %>% 
  select(ancestor) %>% 
  unique() %>% 
  sample_n(100)

pango_short_counts %>%
  arrange(ancestor_lbi) %>% 
  inner_join(node_sample, by = join_by(ancestor)) %>%
  ggplot(aes(x = ancestor, y=pango_short)) +
  geom_col(aes(fill=pango_short, group=ancestor), show.legend = T) +
  scale_x_discrete(
    guide=guide_axis(angle=90), 
    labels = function(s){
      pango_short_counts %>% 
        filter(ancestor %in% s) %>% 
        distinct(ancestor, ancestor_lbi) %>% 
        select(ancestor_lbi) %>% 
        pull %>% 
        round(3)
    }
  )

pango_short_counts %>%
  filter(n_desc_tips >= 1000) %>% 
  ggplot(aes(x=ancestor_lbi, y=n_pango_short/n_desc_tips)) +
  geom_point(aes(color = pango_short), show.legend = T, alpha=0.2)

pango_short_counts %>% 
  mutate(pango_frac = n_pango_short/n_desc_tips) %>% 
  pivot_wider(names_from = pango_short, values_from = pango_frac, names_prefix = 'frac_', values_fill = 0) %>%
  pivot_longer(cols = starts_with('frac_'), names_to = 'pango_short', names_prefix = 'frac_', values_to = 'pango_frac') %>%
  arrange(ancestor_lbi) %>% 
  ggplot(aes(y=pango_short, x=ancestor)) +
  geom_tile(aes(fill=pango_frac)) +
  scale_fill_continuous(type='viridis') +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size=16)) +
  theme(axis.text.y = element_text(size=14)) +
  scale_y_discrete(sec.axis = dup_axis()) +
  scale_x_discrete(
    labels = function(s){
      lbis = pango_short_counts %>%
        arrange(ancestor_lbi) %>%
        filter(ancestor %in% s) %>%
        distinct(ancestor, ancestor_lbi) %>%
        select(ancestor_lbi) %>%
        pull %>%
        round(3)
      for (i in seq(length(lbis))){
        if (i %% 20 != 0) {
          lbis[i] = ''
        }
      }
      return(lbis)
    }
  )

```


