---
title: "h5n1_segment_dms_summary"
output: html_document
date: '`r format(Sys.Date(), "%d %B %Y")`'
---

This file deals with mutations called against the DMS references, rather than against ancestral sequences.

# Setup
```{r setup, include=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggrepel)
library(stringr)
library(ggpubr)
library(ggpmisc)
library(lubridate)
library(ape)
library(ggtree)
library(phangorn)

theme_set(theme_bw())
setwd('/Users/james/Documents/convergent')
```

```{r read_mutations_data, include=F, echo=F}
mutations = read.csv('data/dms_mutations_wrt_ref_avian_ha.tsv', sep = '\t') %>%
  mutate(host = 'avian', gene = 'HA') %>%
  bind_rows(
    read.csv('data/dms_mutations_wrt_ref_bovine_ha.tsv', sep = "\t") %>%
      mutate(host = 'bovine', gene = 'HA')
  ) %>% 
  bind_rows(
    read.csv('data/dms_mutations_wrt_ref_avian_pb2.tsv', sep = "\t") %>%
      mutate(host = 'avian', gene = 'PB2')
  ) %>% 
  bind_rows(
    read.csv('data/dms_mutations_wrt_ref_bovine_pb2.tsv', sep = "\t") %>%
      mutate(host = 'bovine', gene = 'PB2')
  ) %>% 
  rename(
    dms_name = phenotype_metric_name,
    dms_value = value,
    pos_aa = position_aa
  ) %>% 
  mutate(
    date = ymd(date)
  ) %>% 
  filter(dms_name != 'mature_h5_site')
```

```{r set_up_ancestral_mutation_data, include=F, echo=F}
# read ancestral mutations
ancestor_muts_dms = read.csv('out_bovine_b3.13_ha/mutations_wrt_ref.tsv', sep = "\t") %>% 
  mutate(host='bovine', gene = 'HA') %>% 
  bind_rows(
    read.csv('out_avian_b3.13_ha/mutations_wrt_ref.tsv', sep = "\t") %>% 
    mutate(host='avian', gene = 'HA')
  ) %>% 
  bind_rows(
    read.csv('out_bovine_b3.13_pb2/mutations_wrt_ref.tsv', sep = "\t") %>% 
    mutate(host='bovine', gene = 'PB2')
  ) %>% 
  bind_rows(
    read.csv('out_avian_b3.13_pb2/mutations_wrt_ref.tsv', sep = "\t") %>% 
    mutate(host='avian', gene = 'PB2')
  ) %>% 
  rename(accession = node) %>% 
  filter(str_starts(accession, 'NODE_'))

# read dates for ancestors
ancestor_dates = read.csv('out_avian_b3.13_ha/treetime/dates.tsv', sep='\t') %>% 
  mutate(host='avian', gene='HA') %>% 
  select(! c(numeric.date)) %>% 
  bind_rows(
    read.csv('out_bovine_b3.13_ha/treetime/dates.tsv', sep='\t') %>% 
      mutate(host='bovine', gene='HA') %>% 
      select(! c(numeric.date))
  ) %>% 
   bind_rows(
    read.csv('out_avian_b3.13_pb2/treetime/dates.tsv', sep='\t') %>% 
      mutate(host='avian', gene='PB2') %>% 
      select(! c(numeric.date))
  ) %>%
   bind_rows(
    read.csv('out_bovine_b3.13_pb2/treetime/dates.tsv', sep='\t') %>% 
      mutate(host='bovine', gene='PB2') %>% 
      select(! c(numeric.date))
  ) %>% 
  rename(accession = X.node) %>% 
  filter(str_starts(accession, 'NODE_')) %>% 
  mutate(date = ymd(date))

# join in dates
ancestor_muts_dms = left_join(
  ancestor_muts_dms,
  ancestor_dates,
  by = join_by(accession, host, gene)
)

# read dms data
dms = read.csv('data/dms_data_ha.tsv', sep = '\t') %>%
  rename(pos_aa = position_aa,
         dms_name = name,
         dms_value = value,) %>%
  mutate(gene = 'HA') %>%
  select(!c('ref_codon', 'alt_codon', 'gff_feature')) %>%
  filter(!dms_name %in% c('mature_h5_site')) %>%
  distinct() %>%
  bind_rows(
    read.csv('data/summary_prefs_effects_diffsel.csv') %>%
      select(c('site', 'wildtype', 'mutation', 'mutdiffsel')) %>%
      rename(
        pos_aa = site,
        ref_aa = wildtype,
        alt_aa = mutation
      ) %>%
      mutate(gene = 'PB2') %>%
      pivot_longer(
        cols = mutdiffsel,
        names_to = 'dms_name',
        values_to = 'dms_value'
      ) %>%
      filter(!is.na(dms_value))
  )

# join in dms data
ancestor_muts_dms = ancestor_muts_dms %>% 
  left_join(
    dms,
    by = join_by(gene, pos_aa, ref_aa, alt_aa),
    relationship = "many-to-many"
  )

```


```{r sum_dms_values, include=F, echo=F}
mutations_summed_dms =
  mutations %>%
  summarize(
    dms_value_sum = sum(dms_value),
    n_dms_values = n(),
    .by = c(accession, host, date, dms_name, gene)
  ) %>%
  mutate(seq_type = 'sample') %>%
  bind_rows(
    ancestor_muts_dms %>%
      filter(!is.na(dms_name)) %>% 
      summarize(
        dms_value_sum = sum(dms_value),
        n_dms_values = n(),
        .by = c(accession, date, host, dms_name, gene)
      ) %>% 
      mutate(seq_type = 'ancestral')
  )
```

## QC Plots

```{r count_dms_values_summed, include=T, echo=F, fig.width=14, fig.height=9}
# How many values represent each thing
mutations_summed_dms %>% filter(seq_type == 'sample') %>% 
  ggplot(aes(x=n_dms_values)) +
  geom_bar() +
  facet_wrap(vars(dms_name, host, gene), labeller = label_wrap_gen(multi_line=FALSE), scales = 'free_y') +
  ggtitle("count dms values summed: samples")

mutations_summed_dms %>% filter(seq_type == 'ancestral') %>% 
  ggplot(aes(x=n_dms_values)) +
  geom_bar() +
  facet_wrap(vars(dms_name, host, gene), labeller = label_wrap_gen(multi_line=FALSE), scales = 'free_y') +
  ggtitle("count dms values summed: ancestral nodes")
```

### Missing DMS Values: ancestral seqs

How many of our mutations do we have DMS data for?
This is counting by position, ref, and alt. 
This is only for ancestral sequence mutations (wrt ref).
```{r missing_dms_values, include=T, echo=F}
# count how many positions are missing values
wide_dms_by_mutation = ancestor_muts_dms %>% 
  filter(alt_aa != ref_aa) %>% 
  pivot_wider(
    names_from=c(dms_name),
    values_from=dms_value,
    names_prefix='dms_'
  ) %>% 
  # shim is na for all, which lets us get the totals easily
  mutate(dms_shim_const_NA=NA)

calc_fracs_present = function(df){
  return(
    df %>%  
    mutate(across(starts_with('dms_'), ~ 1 - (.x/dms_shim_const_NA), .names='frac_present_{.col}')) %>%
    select(!(starts_with('dms_') | contains('NA', ignore.case=F))) %>% 
    pivot_longer(names_to='dms_name', values_to='frac_present', cols=starts_with('frac_')) %>% 
    mutate(dms_name = str_remove(dms_name, 'frac_present_dms_')) 
    # pivot_wider(names_from = c(host), values_from=frac_present, names_prefix = 'frac_present_')
  )
}

dms_values_present_by_mutation = wide_dms_by_mutation %>% 
  summarize(across(starts_with('dms_'), ~ sum(is.na(.x))), .by=c(host, gene)) %>% 
  calc_fracs_present()

ggplot(dms_values_present_by_mutation, aes(x=dms_name, y=frac_present)) +
  geom_col(aes(fill=interaction(host, gene)), position=position_dodge()) +
  theme(axis.text.x = element_text(angle = -30, vjust = 0.5, hjust=0))
```

# Overall distributions
## Samples

```{r dist_of_summed_values, include=T, echo=F, fig.width=13, fig.height=10}
mutations_summed_dms %>% filter(seq_type == 'sample') %>% 
ggplot(aes(x = dms_value_sum)) +
  geom_histogram() +
  facet_wrap(vars(dms_name, host, gene), labeller = label_wrap_gen(multi_line=FALSE), scales='free', ncol=4) +
  ggtitle('distribution of dms sums within sample seqs')

mutations_summed_dms %>% filter(seq_type == 'sample') %>% 
ggplot(aes(x = date, y=dms_value_sum)) +
  geom_point(aes(color = n_dms_values)) +
  scale_color_viridis_c() +
  facet_wrap(vars(dms_name, host, gene), labeller = label_wrap_gen(multi_line=FALSE), scales='free_y', ncol=4) +
  ggtitle('dms sums by date within sample seqs')
```

## Ancestors
```{r dist_of_summed_values_ancestors, include=T, echo=F, fig.width=13, fig.height=10}
mutations_summed_dms %>% filter(seq_type == 'ancestral') %>% 
ggplot(aes(x = dms_value_sum)) +
  geom_histogram() +
  facet_wrap(vars(dms_name, host, gene), labeller = label_wrap_gen(multi_line=FALSE), scales='free', ncol=4) +
  ggtitle('distribution of dms sums within ancestral seqs')

mutations_summed_dms %>% filter(seq_type == 'ancestral') %>% 
ggplot(aes(x = date, y=dms_value_sum)) +
  geom_point(aes(color = n_dms_values)) +
  scale_color_viridis_c() +
  facet_wrap(vars(dms_name, host, gene), labeller = label_wrap_gen(multi_line=FALSE), scales='free_y', ncol=4) +
    ggtitle('dms sums by date within ancestral seqs')
```

# Go to Ombrosa, take to the trees

```{r set_up_trees, include=F, echo=F}
tree_bovine_ha = read.nexus('out_bovine_b3.13_ha/ancestral/annotated_tree.nexus')
tree_avian_ha = read.nexus('out_avian_b3.13_ha/ancestral/annotated_tree.nexus')
tree_bovine_pb2 = read.nexus('out_bovine_b3.13_pb2/ancestral/annotated_tree.nexus')
tree_avian_pb2 = read.nexus('out_avian_b3.13_pb2/ancestral/annotated_tree.nexus')

get_dms_values = function(nodes, dms_name) {
  joined = data.frame(accession=nodes, dms_name=dms_name) %>% 
    left_join(mutations_summed_dms, by=join_by(accession, dms_name)) %>% 
    select(accession, dms_value_sum)
  return(joined)
}

```

```{r calc_deltas_parent_to_child, include=F, echo=F}
get_parent_child_deltas = function(tree, host_, gene_) {
  edge_dms = data.frame()
  edges = tree$edge %>% 
      as.data.frame() %>% 
      mutate(index = row_number())
  
  all_labels = c(tree$tip.label, tree$node.label)
  for(child in all_labels) {
    parent = all_labels[Ancestors(tree, child, type='parent')]
    if(length(parent) == 0){
      next
    }
    child_idx = which(all_labels == child)
    parent_idx = which(all_labels == parent)
    edge_idx = edges %>% filter(V1 == parent_idx & V2 == child_idx) %>% pull(index)
    
    edge_dms = edge_dms %>% bind_rows(
      mutations_summed_dms %>% 
        filter(gene == gene_ & host == host_ & accession == child) %>%
        inner_join(
          mutations_summed_dms %>% 
              filter(gene == gene_ & host == host_ & accession == parent),
          by = join_by(gene, host, dms_name),
          suffix = c("_child", "_parent")
        ) %>% 
        mutate(edge_index = edge_idx)
    )
  }
  edge_dms = edge_dms %>% 
    mutate(
      dms_value_sum_delta = dms_value_sum_child - dms_value_sum_parent
    )
  return(edge_dms)
}

get_edge_dms_deltas_vec = function(tree, edge_dms, dms_name) {
  n_edges = nrow(tree$edge)
  scaffold = data.frame(edge_index = seq(n_edges), dms_name = dms_name)
  edge_deltas = (
    scaffold %>% 
      left_join(
        edge_dms,
        by = join_by(edge_index, dms_name)
      )
  )$dms_value_sum_delta
  edge_deltas = c(NA, edge_deltas)
  return(edge_deltas)
}

```

## Bovine HA

```{r plot_dms_colored_trees_bovine_ha, include=T, echo=F, fig.height=20, fig.width=10}
plot_dms_tree = function(dms_name, tree, host, gene){
  dms_values = get_dms_values(tree$tip.label, dms_name)
  p = ggtree(tree)
  p = p %<+% dms_values +
      geom_tippoint(aes(color=dms_value_sum), size=2) +
      scale_color_viridis_c() +
      ggtitle(paste(paste(host, gene, "B3.13", sep = ', '), dms_name, sep = " | ")) +
      theme(plot.title = element_text(size = 16, face = "bold", vjust = 0)) + 
      theme(panel.border = element_rect()) +
      expand_limits(y = c(-20, length(tree$tip.label) + 20))
  print(p)
}

for (name in (mutations_summed_dms %>% filter(gene == 'HA') %>% select(dms_name) %>% unique())$dms_name) {
  plot_dms_tree(name, tree_bovine_ha, "Bovine", "HA")
}
```

## Avian HA

```{r plot_dms_colored_trees_avian_ha, include=T, echo=F, fig.height=20, fig.width=10}
for (name in (mutations_summed_dms %>% filter(gene == 'HA') %>% select(dms_name) %>% unique())$dms_name){
  plot_dms_tree(name, tree_avian_ha, 'Avian', "HA")
}
```

## PB2

```{r plot_dms_colored_trees_pb2, include=T, echo=F, fig.height=20, fig.width=10}
plot_dms_tree("mutdiffsel", tree_bovine_pb2, "Bovine", "PB2")
plot_dms_tree("mutdiffsel", tree_avian_pb2, "Avian", "PB2")
```

## Ancestors

```{r plot_dms_colored_trees_ancestors, include=T, echo=F, fig.height=20, fig.width=10}
plot_dms_tree_ancestors = function(dms_name, tree, host, gene){
  dms_values = get_dms_values(tree$node.label, dms_name)
  p = ggtree(tree)
  p = p %<+% dms_values +
      geom_nodepoint(aes(color=dms_value_sum), size=2) +
      scale_color_viridis_c() +
      ggtitle(paste(paste(host, gene, "B3.13", sep = ', '), dms_name, sep = " | ")) +
      theme(plot.title = element_text(size = 16, face = "bold", vjust = 0)) + 
      theme(panel.border = element_rect()) +
      expand_limits(y = c(-20, length(tree$tip.label) + 20))
  print(p)
}

for (name in (mutations_summed_dms %>% filter(gene == 'HA') %>% select(dms_name) %>% unique())$dms_name) {
  plot_dms_tree_ancestors(name, tree_bovine_ha, "Bovine", "HA")
}

for (name in (mutations_summed_dms %>% filter(gene == 'HA') %>% select(dms_name) %>% unique())$dms_name) {
  plot_dms_tree_ancestors(name, tree_avian_ha, "Avian", "HA")
}

plot_dms_tree_ancestors("mutdiffsel", tree_bovine_pb2, "Bovine", "PB2")
plot_dms_tree_ancestors("mutdiffsel", tree_avian_pb2, "Avian", "PB2")
```

### Delta parent-to-child

Here, edges are colored by the change in summed dms value between parent and child.
That is: `child_value - parent_value`.

```{r plot_parent_child_deltas, include=T, echo=F, fig.height=20, fig.width=10}
plot_dms_tree_edge_deltas = function(tree, edge_dms, host, gene, dms_name){
  edge_deltas_vec = get_edge_dms_deltas_vec(tree, edge_dms, dms_name)

    p = ggtree(tree, aes(color=edge_deltas_vec)) +
      scale_color_viridis_c() +
      ggtitle(paste(paste(host, gene, "B3.13", sep = ', '), dms_name, sep = " | ")) +
      labs(color='child - parent') +
      theme(plot.title = element_text(size = 16, face = "bold", vjust = 0)) + 
      theme(panel.border = element_rect()) +
      expand_limits(y = c(-20, length(tree$tip.label) + 20))
  print(p)
}

edge_dms_bovine_ha = get_parent_child_deltas(tree_bovine_ha, 'bovine', 'HA')
for (name in (mutations_summed_dms %>% filter(gene == 'HA') %>% select(dms_name) %>% unique())$dms_name){
  plot_dms_tree_edge_deltas(tree_bovine_ha, edge_dms_bovine_ha, 'bovine', 'HA', name)
}



edge_dms_avian_ha = get_parent_child_deltas(tree_avian_ha, 'avian', 'HA')
for (name in (mutations_summed_dms %>% filter(gene == 'HA') %>% select(dms_name) %>% unique())$dms_name){
  plot_dms_tree_edge_deltas(tree_avian_ha, edge_dms_avian_ha, 'avian', 'HA', name)
}

edge_dms_bovine_pb2 = get_parent_child_deltas(tree_bovine_pb2, 'bovine', 'PB2')
plot_dms_tree_edge_deltas(tree_bovine_pb2, edge_dms_bovine_pb2, 'bovine', 'PB2', 'mutdiffsel')

edge_dms_avian_pb2 = get_parent_child_deltas(tree_avian_pb2, 'avian', 'PB2')
plot_dms_tree_edge_deltas(tree_avian_pb2, edge_dms_avian_pb2, 'avian', 'PB2', 'mutdiffsel')

```


