---
title: "flu_selection"
output: html_document
date: "2026-01-22"
---

```{r setup, include=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggrepel)
library(stringr)

theme_set(theme_bw())
setwd('/Users/james/Documents/convergent')
```
```{r read_files, include=F}
mutations = read.csv('out/codon_mutation_counts_bovine_ha.tsv', sep='\t') %>% 
  mutate(host='bovine')

mutations = full_join(
  mutations,
  read.csv('out_avian_b3.13/codon_mutation_counts_ha.tsv', sep='\t') %>% 
    mutate(host='avian')
)

mutations = mutations %>% 
  group_by(pos_aa, ref_aa, alt_aa, host) %>% 
  summarize(count = sum(count))
```

The DMS value calculation is slightly different than before, but it doesn't make much difference.
As before, we match DMS values with observed mutations by amino acid position, ref, and alt.
But instead of using a simple average across all values for a given position, 
we instead use a weighted average that favors the DMS values associated with mutations with more occurrences.
```{r read_dms, include=F}
dms_ha = read.csv('data/dms_data_ha.tsv', sep='\t') %>% 
  rename(
    pos_aa = position_aa,
    dms_name = name,
    dms_value = value,
  ) %>% 
  select(!c('ref_codon', 'alt_codon', 'gff_feature')) %>% 
  filter(!dms_name %in% c('evescape_sigmoid', 'evescape')) %>% 
  distinct()

# match dms to mutations by amino acid pos, ref, and alt.
# so dms values for mutations not seen in our data are not considered, even if 
# we see other mutations at the same position.
dms_by_mutation = 
  left_join(
    mutations, 
    dms_ha, 
    by=c('pos_aa', 'ref_aa', 'alt_aa'),
    relationship='many-to-many'
  )
  
# Weighted average of DMS values 
# only results in minor changes compared to naive average
dms_by_position = dms_by_mutation %>% 
  filter(!is.na(dms_name)) %>% 
  group_by(pos_aa, dms_name, host) %>% 
  summarize(
    wt_mean_dms_value = sum(count * dms_value)/sum(count),
    mean_dms_value = mean(dms_value),
    min_dms_value = min(dms_value),
    max_dms_value = max(dms_value),
  )

```

```{r missing_dms_values, include=T, echo=F}
# count how many positons are missing values
wide_dms_by_mutation = dms_by_mutation %>% 
  select(!count) %>% 
  filter(alt_aa != ref_aa) %>% 
  pivot_wider(
    names_from=c(dms_name),
    values_from=dms_value,
    names_prefix='dms_'
  ) %>% 
  # shim is na for all, which lets us get the totals easily
  mutate(dms_shim_const_NA=NA)

calc_fracs_present = function(df){
  return(
    df %>%  
    mutate(across(starts_with('dms_'), ~ 1 - (.x/dms_shim_const_NA), .names='frac_present_{.col}')) %>%
    select(!(starts_with('dms_') | contains('NA'))) %>% 
    pivot_longer(names_to='dms_name', values_to='frac_present', cols=starts_with('frac_')) %>% 
    mutate(dms_name = str_remove(dms_name, 'frac_present_dms_')) %>% 
    pivot_wider(names_from = c(host), values_from=frac_present, names_prefix = 'frac_present_')
  )
}

dms_values_present_by_mutation = wide_dms_by_mutation %>% 
  group_by(host) %>% 
  summarize(across(starts_with('dms_'), ~ sum(is.na(.x)))) %>% 
  calc_fracs_present()
print(dms_values_present_by_mutation)

dms_values_present_by_site = wide_dms_by_mutation %>% 
  group_by(host, pos_aa) %>% 
  summarize(across(starts_with('dms_'), ~ mean(.x, na.rm=T))) %>% 
  group_by(host) %>% 
  summarize(across(starts_with('dms_'), ~ sum(is.nan(.x)))) %>% 
  calc_fracs_present()
print(dms_values_present_by_site)
```

```{r calc_dn_ds, include=F}
dn_ds_ha = mutations %>% 
  mutate(synonymous = (alt_aa == ref_aa)) %>% 
  group_by(pos_aa, host, synonymous) %>% 
  summarize(n = sum(count)) %>% 
  pivot_wider(names_from=synonymous, values_from=n, values_fill=0) %>% 
  rename(count_n = 'FALSE', count_s = 'TRUE') %>% 
  mutate(
    ratio = case_when(
      count_s == 0 ~ count_n,
      .default = count_n / count_s
    ),
    zero_synonymous = (count_s == 0)
  )

dn_ds_dms_ha = left_join(dms_by_position, dn_ds_ha, by=c('pos_aa', 'host'))
```


```{r plot_by_pos, include=T, echo=F}
ggplot(dn_ds_ha, aes(x=pos_aa, y=ratio)) +
  geom_point(aes(color=zero_synonymous)) +
  facet_wrap(~host, ncol=1) +
  geom_text_repel(aes(label=ifelse(ratio>2.5,pos_aa,''))) +
  ggtitle('Selection in HA, B3.13') +
  xlab('amino acid position') +
  ylab('dn/ds')
```


```{r plot_dms_comparison, include=T, echo=F, warning=F, fig.height=11, fig.width=11}
ggplot(dn_ds_dms_ha, aes(x=wt_mean_dms_value, y=ratio)) +
  geom_point(aes(color=zero_synonymous)) +
  ggtitle('HA, B3.13') +
  ylab('dn/ds') +
  facet_wrap(vars(dms_name, host), scales='free_x', ncol=4) +
  geom_text_repel(aes(label=pos_aa), max.overlaps=5, size=4)
```

Error bars show the range of DMS values associated with each point.
```{r plot_dms_comparison_error_bars, include=T, echo=F, warning=F, fig.height=11, fig.width=11}
ggplot(dn_ds_dms_ha, aes(x=wt_mean_dms_value, y=ratio, color=pos_aa)) +
  geom_point() +
  scale_color_continuous(type='viridis') + 
  geom_errorbarh(aes(xmin=min_dms_value, xmax=max_dms_value), height=0.5) +
  ggtitle('HA, B3.13') +
  ylab('dn/ds') +
  facet_wrap(vars(dms_name, host), scales='free_x', ncol=4) +
  geom_text_repel(aes(label=pos_aa), color='black', max.overlaps=5, size=4)
```
