---
title: "flu_selection"
output: html_document
date: "2026-01-22"
---
## Setup
```{r setup, include=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggrepel)
library(stringr)
library(ggpubr)
library(ggpmisc)

theme_set(theme_bw())
setwd('/Users/james/Documents/convergent')
```

```{r read_files, include=F}
# HA data
mutations = read.csv('out/codon_mutation_counts_bovine_ha.tsv', sep='\t') %>% 
  mutate(host='bovine')
mutations = full_join(
  mutations,
  read.csv('out_avian_b3.13/codon_mutation_counts_ha.tsv', sep='\t') %>% 
    mutate(host='avian')
)
mutations = mutations %>% mutate(gene='HA')

# PB2 data
mutations = full_join(
  mutations,
  read.csv('out_bovine_b3.13_pb2/codon_mutation_counts.tsv', sep='\t') %>% 
    mutate(host='bovine', gene='PB2')
)
mutations = full_join(
  mutations,
  read.csv('out_avian_b3.13_pb2/codon_mutation_counts.tsv', sep='\t') %>% 
    mutate(host='avian', gene='PB2')
)

mutations = mutations %>% 
  group_by(pos_aa, ref_aa, alt_aa, host, gene) %>% 
  summarize(count = sum(count))

ha_site_numbering = read.csv('data/site_numbering_map_HA.csv')

```

### DMS data

#### Average value calculation

The DMS value calculation is slightly different than before, but it doesn't make much difference.
As before, we match DMS values with observed mutations by amino acid position, ref, and alt.
But instead of using a simple average across all values for a given position, 
we instead use a weighted average that favors the DMS values associated with mutations with more occurrences.

For example, if we have the following mutations:
```{r table1, echo=F}
print(data.frame(
  pos = c(10, '"'),
  ref = c('I', '"'),
  alt = c('M', 'V'),
  host = c('bovine', '"'),
  count = c(2, 1)
))
```
And the following DMS values for a given metric:
```{r table2, echo=F}
print(data.frame(
  pos = c(10, '"'),
  ref = c('I', '"'),
  alt = c('M', 'V'),
  value = c(0.2676, -2.3170)
))
```
Then we calculate our weighted average DMS value for bovine position 10 as:

```
sum(count * dms_value) / sum(count) 
= (2 * 0.2676 + 1 * -2.317) / 3 
= -0.593933
```
#### Read DMS data

todo: numbering for PB2 data

```{r read_dms, include=F}
dms_ha = read.csv('data/dms_data_ha.tsv', sep='\t') %>% 
  rename(
    pos_aa = position_aa,
    dms_name = name,
    dms_value = value,
  ) %>% 
  mutate(gene='HA') %>% 
  select(!c('ref_codon', 'alt_codon', 'gff_feature')) %>% 
  filter(!dms_name %in% c('evescape_sigmoid', 'evescape', 'mature_h5_site')) %>% 
  distinct()

# todo: numbering?
dms_pb2 = read.csv('data/summary_prefs_effects_diffsel.csv') %>% 
  select(c('site', 'wildtype', 'mutation', 'mutdiffsel')) %>% 
  rename(
    pos_aa = site,
    ref_aa = wildtype,
    alt_aa = mutation
  ) %>% 
  mutate(gene='PB2') %>% 
  pivot_longer(cols=mutdiffsel, names_to='dms_name', values_to='dms_value') %>% 
  filter(!is.na(dms_value))

dms = full_join(
  dms_ha,
  dms_pb2
)
  

# match dms to mutations by amino acid pos, ref, and alt.
# so dms values for mutations not seen in our data are not considered, even if 
# we see other mutations at the same position.
dms_by_mutation = 
  left_join(
    mutations, 
    dms, 
    by=c('pos_aa', 'ref_aa', 'alt_aa', 'gene'),
    relationship='many-to-many'
  )
  

```

#### Missing DMS Values

How many of our mutations do we have DMS data for?
This is counting by position, ref, and alt. 
```{r missing_dms_values, include=T, echo=F}
# count how many positons are missing values
wide_dms_by_mutation = dms_by_mutation %>% 
  select(!count) %>% 
  filter(alt_aa != ref_aa) %>% 
  pivot_wider(
    names_from=c(dms_name),
    values_from=dms_value,
    names_prefix='dms_'
  ) %>% 
  # shim is na for all, which lets us get the totals easily
  mutate(dms_shim_const_NA=NA)

calc_fracs_present = function(df){
  return(
    df %>%  
    mutate(across(starts_with('dms_'), ~ 1 - (.x/dms_shim_const_NA), .names='frac_present_{.col}')) %>%
    select(!(starts_with('dms_') | contains('NA', ignore.case=F))) %>% 
    pivot_longer(names_to='dms_name', values_to='frac_present', cols=starts_with('frac_')) %>% 
    mutate(dms_name = str_remove(dms_name, 'frac_present_dms_')) 
    # pivot_wider(names_from = c(host), values_from=frac_present, names_prefix = 'frac_present_')
  )
}

dms_values_present_by_mutation = wide_dms_by_mutation %>% 
  group_by(host, gene) %>% 
  summarize(across(starts_with('dms_'), ~ sum(is.na(.x))), .groups='keep') %>% 
  calc_fracs_present()

ggplot(dms_values_present_by_mutation, aes(x=dms_name, y=frac_present)) +
  geom_col(aes(fill=interaction(host, gene)), position=position_dodge()) +
  theme(axis.text.x = element_text(angle = -30, vjust = 0.5, hjust=0))

dms_values_present_by_site = wide_dms_by_mutation %>%
  group_by(host, gene, pos_aa) %>%
  summarize(across(starts_with('dms_'), ~ mean(.x, na.rm=T))) %>%
  group_by(host, gene) %>%
  summarize(across(starts_with('dms_'), ~ sum(is.nan(.x)))) %>%
  calc_fracs_present()

```

### Calculate dN / dS

```{r calc_dn_ds, include=F}
dn_ds = mutations %>% 
  mutate(synonymous = (alt_aa == ref_aa)) %>% 
  group_by(pos_aa, host, gene, synonymous) %>% 
  summarize(n = sum(count)) %>% 
  pivot_wider(names_from=synonymous, values_from=n, values_fill=0) %>% 
  rename(count_n = 'FALSE', count_s = 'TRUE') %>% 
  mutate(
    ratio_formula = case_when(
      count_s == 0 ~ 'count_n',
      count_n == 0 & count_s >= 1 ~ '1/count_s',
      .default = 'count_n/count_s'
    ),
    ratio = case_when(
      count_s == 0 ~ count_n,
      count_n == 0 & count_s >= 1 ~ 1 / count_s,
      .default = count_n / count_s
    ),
    zero_synonymous = (count_s == 0)
  )

# add region to dn_ds
dn_ds = dn_ds %>% 
  left_join(
    ha_site_numbering %>% 
      select(sequential_site, region) %>% 
      rename(pos_aa = sequential_site) %>% 
      mutate(gene = 'HA'),
    by=c('pos_aa', 'gene')
  )

dn_ds_dms_by_mutation = left_join(
  dms_by_mutation,
  dn_ds,
  by=c('pos_aa', 'host', 'gene'),
  relationship='many-to-many'
) %>% 
  filter(! is.na(dms_name))


dn_ds_dms_by_mutation_ha = dn_ds_dms_by_mutation %>% filter(gene == 'HA')
dn_ds_dms_by_mutation_pb2 = dn_ds_dms_by_mutation %>% filter(gene == 'PB2')
```

## Plots

### dN / dS by position

```{r plot_by_pos, include=T, echo=F, fig.width=12, fig.height=9}
ggplot(dn_ds, aes(x=pos_aa, y=log(ratio))) +
  geom_hline(yintercept=0, linetype='dashed') +
  geom_point(aes(color=ratio_formula)) +
  facet_wrap(vars(gene, host), ncol=1, labeller = label_wrap_gen(multi_line=FALSE)) +
  geom_text_repel(aes(label=ifelse(ratio>2.5,pos_aa,''))) +
  ggtitle('Selection, B3.13') +
  xlab('amino acid position') +
  ylab('log(dN/dS)')
```

```{r praneeths_dn_ds_by_site, include=T, echo=F, warning=F, fig.width=12, fig.height=9}
dn_ds_ha = dn_ds %>% filter(gene == 'HA')
ggplot() +
  geom_rect(data=ha_site_numbering, aes(xmin=sequential_site-0.5, xmax=sequential_site+0.5, ymin=-Inf, ymax=Inf, fill=region), alpha=0.7) +
  scale_fill_manual(values=c('#77AADD', '#AAAA00', '#EE8866', '#99DDFF', '#EEDD88', '#FFAABB', '#44BB99', '#BBCC33')) +
  geom_hline(yintercept=1, linetype='dashed') +
  geom_point(data=dn_ds_ha, aes(x=pos_aa, y=ratio), fill='#0077bb', color='white', shape=21, size=3) +
  facet_wrap(~host, ncol=1) +
  geom_text_repel(data=dn_ds_ha, aes(x=pos_aa, y=ratio, label=ifelse(ratio>2.5,pos_aa,'')), max.overlaps=2) +
  ggtitle('Selection in HA, B3.13') +
  xlab('') +
  ylab('dN/dS Ratio') +
  scale_x_continuous(breaks=seq(0, 700, by=100)) +
  theme_classic()

```

### DMS comparison
#### HA

Plot with one point per observed mutation:

```{r plot_dms_comparison_ha, include=T, echo=F, warning=F, fig.height=11, fig.width=11}
ggplot(dn_ds_dms_by_mutation_ha, aes(x=dms_value, y=log(ratio))) +
  geom_hline(yintercept = 0, linetype='dashed') +
  geom_point(aes(color=count)) +
  scale_color_continuous(type='viridis') +
  ggtitle('HA, B3.13') +
  ylab('log(dN/dS)') +
  facet_wrap(vars(dms_name, host), scales='free_x', ncol=4) +
  geom_text_repel(aes(label=paste0(ref_aa, pos_aa, alt_aa)), max.overlaps=2, size=3, box.padding=0)
  
```

#### PB2
By mutation:

```{r plot_dms_comparison_pb2, include=T, echo=F, warning=F, fig.height=5, fig.width=11}
ggplot(dn_ds_dms_by_mutation_pb2, aes(x=dms_value, y=log(ratio))) +
  geom_hline(yintercept = 0, linetype='dashed') +
  geom_point(aes(color=count)) +
  scale_color_continuous(type='viridis') +
  ggtitle('PB2, B3.13') +
  ylab('log(dN/dS)') +
  facet_wrap(vars(dms_name, host), scales='free_x', ncol=4) +
  geom_text_repel(aes(label=paste0(ref_aa, pos_aa, alt_aa)), max.overlaps=5, size=4)
```


### Sites of interest selection

```{r HA_trailer_regions, include=T, echo=F, warning=F}
foo = dn_ds_ha %>% filter(region != 'HA1' & host == 'bovine')

hi = foo %>% ungroup %>%  slice_max(ratio, n=5)
lo = foo %>% ungroup %>%  slice_min(ratio, n=5)
extreme = hi %>% union_all(lo)

ggplot(foo, aes(x=pos_aa, y=log(ratio))) +
  scale_color_manual(values=c('#77AADD', '#AAAA00', '#EE8866', '#99DDFF', '#EEDD88', '#FFAABB', '#44BB99', '#BBCC33')[-1]) +
  geom_hline(yintercept=0, linetype='dashed') +
  geom_point(aes(color=region)) +
  facet_wrap(vars(gene, host), ncol=1, labeller = label_wrap_gen(multi_line=FALSE)) +
  geom_text_repel(aes(label=ifelse(pos_aa %in% extreme$pos_aa, pos_aa, '')), size=3, max.overlaps=2, box.padding = 0) +
  ggtitle('B3.13, bovine only, HA1 sites dropped') +
  xlab('amino acid position') +
  ylab('log(dN/dS)')
```

```{r, HA_mouse_ferret, include=T, echo=F, warning=F}
foo = dn_ds_dms_by_mutation_ha %>% 
  filter(region != 'HA1' & host == 'bovine' & dms_name %in% c('mouse_sera_escape', 'ferret_sera_escape'))

ggplot(foo, aes(x=dms_value, y=log(ratio))) +
  geom_hline(yintercept = 0, linetype='dashed') +
  geom_point(aes(color=count)) +
  scale_color_continuous(type='viridis') +
  ggtitle('HA, B3.13, Bovine, HA1 sites dropped') +
  ylab('log(dN/dS)') +
  facet_wrap(vars(dms_name), scales='free_x', ncol=4) +
  geom_text_repel(aes(label=paste0(ref_aa, pos_aa, alt_aa)), max.overlaps=5, size=3, box.padding = 0)
```


```{r HA_trailer_regions_grid, include=T, echo=F, warning=F}
foo = dn_ds_ha %>% 
  rename(subregion = region) %>% 
  mutate(
    region = case_when(
      str_detect(subregion, 'region-[A-E]') ~ 'HA1',
      .default=subregion
    ),
    selection_type = case_when(
      ratio == 1 ~ 'neutral',
      ratio > 1 ~ 'positive',
      ratio < 1 ~ 'purifying',
      .default = 'unset'
    )
  )

hi = foo %>% ungroup %>% slice_max(ratio, n=5, by=host)
lo = foo %>% ungroup %>% slice_min(ratio, n=5, by=host)
extreme = hi %>% union_all(lo) %>% 
  mutate(label=T) %>% 
  select(pos_aa, host, label)

foo = foo %>% 
  left_join(extreme, by=join_by(host, pos_aa)) %>% 
  mutate(
    label = case_when(
      is.na(label) ~ F,
      .default=label
    )
  )

bovine = foo %>% filter(host == 'bovine')
ylim_bovine = log(c(min(foo$ratio), max(foo$ratio)))

p1 = ggplot(foo %>% filter(region == 'HA1'), aes(x=pos_aa, y=log(ratio)))+
  geom_hline(yintercept=0, linetype='dashed') +
  geom_point(aes(color=selection_type)) +
  scale_color_manual(values=c('#a3a3a388', '#66ccee', '#ee6677'))+
  ylim(ylim_bovine) +
  xlab('') +
      ggtitle('B3.13') +
  ylab('log(dN/dS)') +
  theme(
    plot.margin = margin(0, 0, 0, 0)
  )+
  facet_wrap(
    vars(host, region), 
    ncol=1, 
    labeller = label_wrap_gen(multi_line=FALSE),
    scales='fixed',
  )+
  geom_text_repel(aes(label=ifelse(label & ratio < 1, pos_aa, '')), size=3, max.overlaps=3, box.padding = 0) +
    geom_text_repel(aes(label=ifelse(label & ratio > 1 & count_s > 0, pos_aa, '')), size=3, max.overlaps=3, box.padding = 0) +
    geom_text_repel(aes(label=ifelse(label & ratio > 1 & count_s == 0, pos_aa, '')), size=3, max.overlaps=3, box.padding = 0, color='red')
  

p2 = ggplot(foo %>% filter(region == 'PBCS'), aes(x=pos_aa, y=log(ratio)))+
  geom_hline(yintercept=0, linetype='dashed') +
  geom_point(aes(color=selection_type)) +
  scale_color_manual(values=c('#a3a3a388', '#66ccee', '#ee6677'))+
  ylim(ylim_bovine) +
  xlab('amino acid position') +
  theme(
    axis.title.y=element_blank(),
    axis.line.y=element_blank(),
    axis.ticks.y=element_blank(),
    axis.text.y = element_blank(),
    plot.margin = margin(0, 0, 0, 0)
  ) +
  facet_wrap(
    vars(host, region), 
    ncol=1, 
    labeller = label_wrap_gen(multi_line=FALSE),
    scales='fixed',
  )+
  geom_text_repel(aes(label=ifelse(label & ratio < 1, pos_aa, '')), size=3, max.overlaps=3, box.padding = 0) +
    geom_text_repel(aes(label=ifelse(label & ratio > 1 & count_s > 0, pos_aa, '')), size=3, max.overlaps=3, box.padding = 0) +
    geom_text_repel(aes(label=ifelse(label & ratio > 1 & count_s == 0, pos_aa, '')), size=3, max.overlaps=3, box.padding = 0, color='red')

p3 = ggplot(foo %>% filter(region == 'HA2'), aes(x=pos_aa, y=log(ratio)))+
  geom_hline(yintercept=0, linetype='dashed') +
  geom_point(aes(color=selection_type)) +
  scale_color_manual(values=c('#a3a3a388', '#66ccee', '#ee6677'))+
  ylim(ylim_bovine) +
  xlab('') +
  theme(
    axis.title.y=element_blank(),
    axis.line.y=element_blank(),
    axis.ticks.y=element_blank(),
    axis.text.y = element_blank(),
    plot.margin = margin(0, 0, 0, 0)
  )+
  facet_wrap(
    vars(host, region), 
    ncol=1, 
    labeller = label_wrap_gen(multi_line=FALSE),
    scales='fixed',
  )+
  geom_text_repel(aes(label=ifelse(label & ratio < 1, pos_aa, '')), size=3, max.overlaps=3, box.padding = 0) +
    geom_text_repel(aes(label=ifelse(label & ratio > 1 & count_s > 0, pos_aa, '')), size=3, max.overlaps=3, box.padding = 0) +
    geom_text_repel(aes(label=ifelse(label & ratio > 1 & count_s == 0, pos_aa, '')), size=3, max.overlaps=3, box.padding = 0, color='red')

ggarrange(
  p1, p2, p3,
  nrow=1,
  common.legend=T,
  legend='bottom',
  align='h',
  widths = c(5, 2.1, 3)
)

```

## Counts

```{r count_by_position, include=T, echo=F, warning=F, fig.width=12, fig.height=9}
muts = mutations %>% 
  mutate(
    type = case_when(
      ref_aa == alt_aa ~ 'synonymous',
      .default = 'non-synonymous'
    )
  )

ggplot(muts, aes(x=pos_aa, y=count)) +
  geom_point(aes(color=type)) +
  facet_wrap(vars(gene, type, host), ncol=2, labeller = label_wrap_gen(multi_line=FALSE)) +
  geom_text_repel(aes(label=paste0(ref_aa, pos_aa, alt_aa)), size=3, max.overlaps=3, box.padding = 0) +
  ggtitle('B3.13') +
  xlab('amino acid position') +
  ylab('Count of emergences')
```

```{r plot_count_vs_dms_ha, include=T, echo=F, warning=F, fig.height=10, fig.width=11}
ggplot(dn_ds_dms_by_mutation_ha, aes(x=dms_value, y=count)) +
  geom_point() +
  ggtitle('HA, B3.13') +
  ylab('emergence count') +
  facet_wrap(vars(dms_name, host), scales='free_x', ncol=4, labeller = label_wrap_gen(multi_line=FALSE)) +
  geom_text_repel(aes(label=paste0(ref_aa, pos_aa, alt_aa)), max.overlaps=3, size=3, box.padding = 0)
```

```{r plot_count_vs_dms_pb2, include=T, echo=F, warning=F, fig.height=5, fig.width=11}
ggplot(dn_ds_dms_by_mutation_pb2, aes(x=dms_value, y=count)) +
  geom_point() +
  ggtitle('PB2, B3.13') +
  ylab('emergence count') +
  facet_wrap(vars(dms_name, host), scales='free_x', ncol=4) +
  geom_text_repel(aes(label=paste0(ref_aa, pos_aa, alt_aa)), max.overlaps=5, size=4, box.padding = 0)
```

The relationship between count and dN/dS is weak at best.
```{r plot_count_vs_dn_ds, include=T, echo=F, warning=F, fig.height=5, fig.width=5}
ggplot(dn_ds_dms_by_mutation, aes(x=count, y=ratio)) +
  geom_point(alpha=0.1, position = position_jitter(width = 0.2)) +
  stat_poly_line(method = 'lm', se=F)+
  stat_poly_eq() +
  ggtitle('B3.13, non-syn mutations with dms data only') +
  ylab('dN/dS') +
  facet_wrap(vars(host, gene), scales='fixed', ncol=2) 
  # geom_text_repel(aes(label=paste0(ref_aa, pos_aa, alt_aa)), max.overlaps=3, size=4, box.padding = 0)
```

## Flumut

```{r read_flumut, include=F, echo=F}
flumut = read.csv('data/flumut_annotations.tsv', sep='\t')
flumut = flumut %>% 
  rename(
    flumut_effect_name = effect_name
  ) %>% 
  separate_wider_regex(
    mutation_name, 
    patterns=c(
      'gene' = '^.*:',
      'ref_aa' = '\\w',
      'pos_aa' = '\\d+',
      'alt_aa' = '\\w'
      ), 
    ) %>% 
  mutate(
    gene = str_remove(gene, ':'),
    gene = str_remove(gene, '\\d-5'),
    pos_aa = as.numeric(pos_aa) 
  ) %>% 
  filter(gene %in% c('HA', 'PB2')) %>% 
  select(!c(title, authors, year)) %>% 
  unique() %>% 
  # Convert from mature H5 to sequential numbering
  mutate(mature_H5_site = pos_aa) %>% 
  left_join(
    ha_site_numbering %>% select(sequential_site, mature_H5_site) %>% mutate(gene = 'HA'),
    by = join_by(gene, mature_H5_site)
  ) %>% 
  mutate(
    pos_aa = case_when(
      gene == 'HA' ~ sequential_site,
      .default = pos_aa
    )
  ) %>% 
  select(!c(mature_H5_site, sequential_site))
  
flumut_dn_ds_dms = dn_ds_dms_by_mutation %>% 
  left_join(flumut, by=join_by(gene, pos_aa, alt_aa, ref_aa), relationship = 'many-to-many')

flumut_dn_ds = dn_ds %>% 
  left_join(mutations, by=join_by(host, gene, pos_aa), relationship = 'one-to-many') %>% 
  left_join(
    flumut,
    by = join_by(gene, pos_aa, alt_aa, ref_aa),
    relationship = 'many-to-many'
  )
```
The matching between mutations in this dataset and those annotated in flumut is poor.
The lack of significant overlap makes these comparisons mostly useless.
```{r flumut_matching, include=T, echo=F}
flumut_dn_ds %>% 
  mutate(has_flumut = is.na(flumut_effect_name)) %>% 
  mutate(flumut_shim_NA = NA) %>% 
  group_by(gene, host) %>% 
  summarize(across(starts_with('flumut_'), ~ sum(is.na(.x))), .groups='keep') %>% 
  mutate(
    total_muts = flumut_shim_NA,
    count_flumut_present = total_muts - flumut_effect_name
  ) %>% 
  mutate(across(starts_with('flumut_'), ~ 1 - (.x/flumut_shim_NA), .names='frac_present_{.col}')) %>% 
  select(!(starts_with('flumut_') | contains('NA', ignore.case=F))) %>% 
  rename(frac_muts_with_flumut = frac_present_flumut_effect_name) %>% 
  print()
```

```{r plot_flumut_dn_ds, include=T, echo=F, fig.width=10}
ggplot(flumut_dn_ds_dms, aes(x=flumut_effect_name, y=ratio)) +
  geom_hline(yintercept = 1, linetype='dashed')+
  geom_point(aes(color=flumut_effect_name), position = position_jitter(width=0.3), alpha=0.5, show.legend = F) +
  geom_boxplot(outliers=F, alpha=0) +
  coord_flip() +
  ylab('dN/dS') +
  facet_wrap(vars(host, gene), labeller = label_wrap_gen(multi_line=FALSE))

flumut_dn_ds %>% arrange(desc(is.na(flumut_effect_name))) %>% 
  ggplot(aes(x=pos_aa, y=log(ratio))) +
  geom_hline(yintercept=0, linetype='dashed') +
  geom_point(aes(color=flumut_effect_name), alpha=0.4) +
  facet_wrap(vars(gene, host), ncol=1, labeller = label_wrap_gen(multi_line=FALSE)) +
  # geom_text_repel(aes(label=ifelse(ratio>2.5,pos_aa,''))) +
  ggtitle('Selection, B3.13') +
  xlab('amino acid position') +
  ylab('log(dN/dS)')
```

```{r plot_flumut_dms_HA, include=T, echo=F, fig.width=12, fig.height=6}
flumut_dn_ds_dms %>% filter(gene == 'HA' & host == 'bovine') %>% 
ggplot(aes(x=flumut_effect_name, y=dms_value)) +
  geom_point(aes(color=flumut_effect_name), position = position_jitter(width=0.3), alpha=0.5, show.legend = F) +
  geom_boxplot(outliers=F, alpha=0) +
  coord_flip() +
  facet_wrap(vars(host, dms_name), labeller = label_wrap_gen(multi_line=FALSE), scales='free_x') 

flumut_dn_ds_dms %>% 
  filter(gene == 'HA' & host == 'bovine') %>% 
  arrange(desc(is.na(flumut_effect_name))) %>% 
ggplot(aes(x=dms_value, y=log(ratio))) +
  geom_hline(yintercept = 0, linetype='dashed') +
  geom_point(aes(color=flumut_effect_name), alpha=0.4) +
  ggtitle('HA, B3.13') +
  ylab('log(dN/dS)') +
  facet_wrap(vars(dms_name, host), scales='free_x', ncol=4, labeller = label_wrap_gen(multi_line=FALSE)) +
  geom_text_repel(
    aes(
      label=ifelse(
        !is.na(flumut_effect_name), 
        paste0(ref_aa, pos_aa, alt_aa), 
        ''
      ), 
      color=flumut_effect_name
    ), 
    size=3, 
    max.overlaps = Inf
  )
```

```{r plot_flumut_dms_pb2, include=T, echo=F, fig.width = 12, fig.height=6}
flumut_dn_ds_dms %>% filter(gene == 'PB2') %>% 
ggplot(aes(x=flumut_effect_name, y=dms_value)) +
  geom_point(aes(color=flumut_effect_name), position = position_jitter(width=0.3), alpha=0.5, show.legend = F) +
  geom_boxplot(outliers=F, alpha=0) +
  coord_flip() +
  facet_wrap(vars(host, dms_name), labeller = label_wrap_gen(multi_line=FALSE))

flumut_dn_ds_dms %>% 
  filter(gene == 'PB2') %>% 
  arrange(desc(is.na(flumut_effect_name))) %>% 
ggplot(aes(x=dms_value, y=log(ratio))) +
  geom_hline(yintercept = 0, linetype='dashed') +
  geom_point(aes(color=flumut_effect_name), alpha=0.4) +
  ggtitle('PB2, B3.13') +
  ylab('log(dN/dS)') +
  facet_wrap(vars(dms_name, host), scales='free_x', labeller = label_wrap_gen(multi_line=FALSE)) +
  geom_text_repel(
    data = flumut_dn_ds_dms %>% filter(gene == 'PB2' & !is.na(flumut_effect_name)) %>% select(host, pos_aa, ref_aa, alt_aa, dms_value, ratio) %>% unique(),
    aes(
      label = paste0(ref_aa, pos_aa, alt_aa), 
    ), 
    size=3, 
    max.overlaps = 8,
    box.padding = 1,
    nudge_y = 0.1
  )
```




